On the Gravitational stability of gravito-turbulent accretion disks

### 2016-01-28 paper/paper3d.tex
revision 3 by MKL

\begin{abstract}
Low mass, self-gravitating accretion disks admit quasi-steady,
`gravito-turbulent' states in which cooling balances turbulent {\bf viscous?}
heating. However, numerical simulations show that gravito-turbulence
cannot be sustained beyond dynamical timescales when the cooling rate is too fast,
or corresponding turbulent viscosity is too large. The result is is disk fragmentation.        
We motivate and quantify an interpretation of disk % a physical
                                % interpretation 
fragmentation as the inability to maintain gravito-turbulence due to  
{\bf proper / formal / etc?} secondary instabilities driven by:  
\begin{inparaenum}[1)] 
\item 
  cooling, which reduces pressure support; and/or
\item 
  viscosity, which reduces rotational support. 
\end{inparaenum}
We analyze the gravitational stability of accretion disks 
including these non-ideal effects explicitly.    
We derive a general dispersion relation for linear waves in a 
self-gravitating, viscous, non-adiabatic disk with internal heating,
external irradiation, and cooling. We  
consider parameterized cooling functions in two and three dimensions,
as well as %full 
radiative diffusion in 3D. We show that 
%cooling/viscosity can destabilize the disk even when the classic
%Toomre stability condition is satisfied; 
generally there is no critical cooling rate/viscosity below which 
the disk is formally stable, although interesting limits appear when we examine the growth
of modes of order the disk scale height. 
We apply this new linear theory to 
global models of protoplanetary disks (PPDs) subject to
gravito-turbulence modeled as an effective 
viscosity, and cooling via dust opacity. 
%We find the sole effect of
%cooling is insufficient to 
%destabilize a realistic PPD. 
We find that viscosity renders the PPD beyond $\sim 60$AU dynamically
unstable on radial %this is for Mdot = 1d-6  
lengthscales a few times the local disk thickness. This is coincident 
with the empirical condition for disk fragmentation based on a
maximum sustainable stress. 
We suggest turbulent stresses can play an active role in realistic disk 
fragmentation by removing rotational stabilization against 
self-gravity, and that the observed transition in behavior from gravito-turbulent to fragmenting may derive from
instability of the gravito-turbulent state itself.
\end{abstract}

@@

The standard metric for the (inverse) strength of disk self-gravity is
the \emph{Toomre parameter}, 
\begin{align}\label{toomreQ_criterion}
  Q \equiv \frac{c_s\kappa}{\pi G \Sigma}  
\end{align}
\citep{toomre64}. Here, $c_s$ is the sound-speed,
$\kappa$ is the epicyclic frequency (which equals the rotation
frequency $\Omega$ in a Keplerian disk), $\Sigma$ is the surface
density and $G$ is the gravitational constant. The Toomre parameter is a
measure of the destabilizing effect of self-gravity ($G\Sigma$) against 
the stabilzing effect of rotation ($\kappa$) and pressure 
($c_s$). This is evident from the dispersion relation, 
\begin{align}\label{classic_dr}
  s^2 =  2\pi G\Sigma|k| - \kappa^2 - c_s^2k^2,   
\end{align}
which relates the growth rate $s$ and radial wavenumber $k$ for local,
axisymmetric waves in a two-dimensional (2D, razor-thin), inviscid and
adiabatic (including isothermal) disk. Such disturbances are formally
unstable when $Q<1$. Non-axisymmetric modes can develop for larger,
but still order-unity values of $Q$
\citep{lau_bertin78,papaloizou89,papaloizou91}. 
%strictly speaking non axi *local* modes are stable -> only transient
%growth
%true non-axi instability needs global disk  model 
%non-axi GI at slightly larger Q, but still characterized by Q

We emphasize that the oft-used dispersion relation and corresponding Toomre parameter, %It is important to remember that the popular dispersion relation, 
(Eqs. \ref{toomreQ_criterion} and \ref{classic_dr}) are derived from \emph{idealized} conditions: 
the base disk is laminar and does not experience any net thermal
losses\footnote{Isothermal disks implicitly assume  
  heating and thermal losses are exactly balanced at all 
  times, so these do not cool in the current context.} 
(hereafter `cooling').  
However, real disks can cool, for example, %radiative, convective cooling
due to radiative losses. Accretion disks may also be turbulent, the dynamic and thermodynamic effect of which is often modeled through an
effective viscosity \citep{shakura73,lin87,armitage01,rafikov15}. 
%{\bf  more refs here?} 
%in terms of AM transport and heating 
%assess GI in general

Despite this, the Toomre parameter %and its associated
%dispersion relation are
 is still widely applied to viscous (turbulent), cooling
accretion disk models \citep[e.g.][]{gammie01,cossins09, kimura12}, {\bf more refs?}
%%KMK
despite the mismatch in the included underlying physics.
%even though Eq. \ref{toomreQ_criterion}---\ref{classic_dr} were 
%derived from inviscid disks without cooling. 
%e.g. using classic DR to determine what scales need to be resolved

We show in this work that including non-ideal
physics, such as cooling and viscosity, modifies the classic dispersion
relation, and hence the condition for GI. Thus the Toomre condition
alone is insufficient to assess GI in realistic disks.  The goal of
this work is to generalize the analytic treatment of disk GI to allow
a self-consistent discussion of GI in realistic accretion disk
models. Specifically, we will examine how  cooling and viscosity can
lead to GI even when $Q>1$, by removing pressure and rotational
support, respectively.  We also review the origins for the transition from self-regulated GI 
to fragmentation seen in numerical simulations. {\bf add sentence here more like s 1.4 laying
out what you are about to talk about. at the very least a transition sentence or two?}

\subsection{Cooling-driven gravitational instability}\label{cool_gi}
Cooling reduces pressure support against self-gravity. The
cooling time  
$\tcool$ is the timescale over which the disk temperature $T$ is relaxed
to some floor value $T_0$, which may be zero. 
%That is, $\p_t T =
%-(T-T_0)/\tcool$. 
 It is often written as 
\begin{align} \label{beta_def}
  \tcool = \beta \Omega^{-1},
\end{align}
where $\beta$ is the corresponding dimensionless cooling time. This
type of parameterized cooling, first applied by \cite{gammie01},
allows a range of thermodynamic responses to be explored. In reality, PPD cooling is controlled by radiation 
from dust grains \citep{bell94,dalessio97,chiang97}. Although most PPDs subject to GI are optically thick, 
cooling parameterized by the $\beta$ model above only captures optically thin cooling when used in numerical
simulations.
%{\bf physical origin of cooling in PPDs}

%%KMK 
Previous work has quantified the role of cooling primarily as a means to reduce the sound speed term
in $Q$ (though see e.g. \citealt{clarke07}). Here we will quantify how cooling enables GI
even when $Q>1$. For example, if perturbations can
cool towards zero temperature, we find the above dispersion relation is
modified to read 
\begin{align}\label{cool_dr}
s^2 = 2\pi G\Sigma |k| - \kappa^2 - \gamma \left(\frac{\tcool s}{1 +
  \tcool s}\right)c_s^2k^2,
\end{align}
where $\gamma$ is the adiabatic index. Cooling increases the growth
rate by reducing the magnitude of the 
pressure term. In fact, for $t_c\in[0,\infty)$ a formal condition
for instability is  
\begin{align*}
  2\pi G \Sigma |k| > \kappa^2,
\end{align*}
which is what would be obtained from Eq. \ref{classic_dr} with 
pressure neglected ($c_s\to 0$). This condition does not actually
depend on the cooling rate, and instability is
possible for any finite $Q$. %at sufficiently small scales 
This shows that the mere presence of
cooling changes the qualitative nature of GI compared to the simple
Toomre condition.   
%{\bf numerical implications? if using standard beta cooling, then as
%  long as it's there then sufficiently small scales will be
 % unstbale. can access those scales with high resolution. defer to
 % discussion section}

@@

%turbulence <-> viscosity assumption 
For gaseous accretion disks, a Navier-Stokes viscosity is often
implemented (as above) to mimic hydrodynamic or magneto-hydrodynamic
(MHD) turbulence \citep{shakura73}. How well turbulence can be modeled
as an effective viscosity is a separate issue 
\citep{balbus99}. However, it is reasonable to assume that these mechanisms, which are observed in simulations 
to transport angular momentum outwards (and mass inwards),
frustate rotational support. %KMK (This is, after all, the physical mechanism to turbulent mass accretion.) 
%%KMK this comment seems out of place here 
%For example, \cite{fromang05} found numerical evidence that MHD
%turbulence can enhance GI and lead to disk fragmentation (see also
%\S\ref{MHD}).

In this work, we use viscosity to model two possible
physical effects of turbulence: heating via dissipation and angular momentum transport. 
%KMK
We emphasize that our model for GI in the presence of viscosity (hearafter Viscous GI)
does not assume a particular origin for the viscosity. GI itself or other forms of magnetic or hydrodynamic turbulence are allowed
in this framework. We will
generalize the theory of viscous GI to include an energy equation  with viscous
heating, irradiation, explicit cooling, as well as
three-dimensionality (3D) in order to consider viscous GI in
more realistic PPD models.

@@

\subsection{Relevance to gravito-turbulent disk fragmentation}
We develop a general framework for viscous disks without  
assuming a specific origin of the turbulence that this viscosity 
represents. %without assuming the nature of the turbulence represented
            %by this viscosity  
We will, however, apply the theory to `gravito-turbulent'
disks in which the viscosity is associated with some underlying (classic)
GI, as described below.

Consider an initially laminar, $Q>>1$ disk, without an external heating mechanism, as it
cools. The disk temperature will decline until $Q=O(1)$, whence
non-axisymmetric modes grow and heat the disk through the dissipation
of spiral shocks \citep{cossins09}. %new rafikov 16 
This setup permits a quasi-steady, turbulent state with
$Q =O(1)$ in which cooling is balanced by shock heating to maintain
thermal equilibrium \citep{gammie01}. 
One may represent the dynamic and thermodynamic effect of this turbulence by a
viscosity. Then the $\alpha$ and $\beta$ parameters defined above will
be inversely related ({\bf KMK see foreward ref eq...?}).

Numerical experiments, however, show that if the cooling rate or
viscosity is too large, say,   
\begin{align}\label{frag_cond}
  \beta < \beta_\mathrm{c} \quad
  \left(\alpha>\alpha_\mathrm{c}\right), 
\end{align}
then the disk fragments \citep{gammie01,rice05,rice11}. In fact, in
the absence of global effects, numerical simulations show that there are
two --- and only two --- possible outcomes for self-gravitating,
cooling disks: gravito-turbulence or fragmentation. %more accurately: given a total simulation time, there are two
%outcomes 
%sufficiently long sims -> always frag?

There is considerable debate on the exact value of $\beta_\mathrm{c}$ 
and whether or not a critical cooling time can be defined at all 
\citep{meru11,lodato11,meru12,paardekooper12,hopkins13}. {\bf KMK We defer a discussion of
numerical convergence issues to Section \ref{sec-disc}.?}
%mention numerical effects? 
However, it is generally accepted that \emph{steady, gravito-turbulent
  disks do not  
  exist for sufficiently rapid cooling or large viscosity}.  
This  is intriguing because, as highlighted in
\S\ref{cool_gi}---\ref{visc_gi}, these effects can reduce
gravitational stability independently of their influence on the exact value of the classic $Q$. This motivates a physical    
interpretation of disk fragmentation as the inability to maintain a
gravito-turbulent state due to %the action of 
%KMK
secondary instabilities driven by cooling and/or viscosity.

Indeed, we find that the region where a gravito-turbulent PPD is
dynamically unstable agrees with the current consensus on  where, and
the conditions under which, a PPD is expected to fragment based on
Eq. \ref{frag_cond}, with typical critical values cited in the
literature \citep[e.g. $\beta_c\sim
  3,\,\alpha_c\sim0.1$;][]{gammie01,rice05}.

\subsection{Plan}
%plan of this paper 
%aim to analyse GI in realistic disks with minimal input from numerical
%simulation

>>>>

\begin{abstract}
  Low mass, self-gravitating accretion disks admit quasi-steady,
  `gravito-turbulent' states in which cooling balances turbulent viscous 
  heating. However, numerical simulations show that gravito-turbulence
  cannot be sustained beyond dynamical timescales when the cooling rate is too fast,
  or when the corresponding turbulent viscosity is too large. The result is is disk fragmentation.         
  We motivate and quantify an interpretation of disk 
  fragmentation as the inability to maintain gravito-turbulence due to  
  formal secondary instabilities driven by:  
\begin{inparaenum}[1)] 
\item 
  cooling, which reduces pressure support; and/or
\item 
  viscosity, which reduces rotational support. 
\end{inparaenum}
We analyze the gravitational stability of accretion disks 
including these non-ideal effects explicitly.    
We derive a general dispersion relation for linear waves in a 
self-gravitating, viscous, non-adiabatic disk with internal heating,
external irradiation, and cooling. We  
consider parameterized cooling functions in two and three dimensions,
as well as 
radiative diffusion in 3D. We show that 
generally there is no critical cooling rate/viscosity below which 
the disk is formally stable, although interesting limits appear 
for unstable modes with lengthscales on the order of the disk scale height.   
%when we examine the growth 
%of modes of order the disk scale height. 
We apply this new linear theory to 
global models of protoplanetary disks (PPDs) subject to
gravito-turbulence modeled as an effective 
viscosity, and cooling via dust opacity. 
We find that viscosity renders the PPD beyond $\sim 60$AU dynamically
unstable on radial %this is for Mdot = 1d-6  
lengthscales a few times the local disk thickness. This is coincident 
with the empirical condition for disk fragmentation based on a
maximum sustainable stress. 
We suggest turbulent stresses can play an active role in realistic disk 
fragmentation by removing rotational stabilization against 
self-gravity, and that the observed transition in behavior from 
gravito-turbulent to fragmenting may derive from  
instability of the gravito-turbulent state itself.
\end{abstract}

@@

The standard metric for the (inverse) strength of disk self-gravity is
the \emph{Toomre parameter}, 
\begin{align}\label{toomreQ_criterion}
  Q \equiv \frac{c_s\kappa}{\pi G \Sigma}  
\end{align}
\citep{toomre64}. Here, $c_s$ is the sound-speed,
$\kappa$ is the epicyclic frequency (which equals the rotation
frequency $\Omega$ in a Keplerian disk), $\Sigma$ is the surface
density and $G$ is the gravitational constant. The Toomre parameter is a
measure of the destabilizing effect of self-gravity ($G\Sigma$) against 
the stabilzing effect of rotation ($\kappa$) and pressure 
($c_s$). This is evident from the dispersion relation, 
\begin{align}\label{classic_dr}
  s^2 =  2\pi G\Sigma|k| - \kappa^2 - c_s^2k^2,   
\end{align}
which relates the growth rate $s$ and radial wavenumber $k$ for local,
axisymmetric waves in a two-dimensional (2D, razor-thin), inviscid and
adiabatic (including isothermal) disk. 
%Such disturbances are formally
%unstable when $Q<1$.  
When $Q<1$, there is a range of $k$ for which such disturbances are
unstable.  
Non-axisymmetric modes can develop for larger, but still order-unity
values of $Q$ 
\citep{lau78,papaloizou89,papaloizou91}. 
%strictly speaking non axi *local* modes are stable -> only transient
%growth
%true non-axi instability needs global disk  model 
%non-axi GI at slightly larger Q, but still characterized by Q

We emphasize that the oft-used dispersion relation and corresponding
Toomre parameter (Eqs. \ref{toomreQ_criterion} and \ref{classic_dr})
are derived from \emph{idealized} conditions:  
the base disk is laminar and does not experience any net thermal 
losses\footnote{Isothermal disks implicitly assume  
  heating and thermal losses are exactly balanced at all 
  times, so these do not cool in the current context.} 
(hereafter `cooling').  
However, real disks can cool, for example, %radiative, convective cooling
due to radiative losses. Accretion disks may also be turbulent, the 
dynamic and thermodynamic effect of which is often modeled through an 
effective viscosity \citep{shakura73,lin87,armitage01,rafikov15}. 
%{\bf  more refs here?} 
%in terms of AM transport and heating 
%assess GI in general

However, the Toomre parameter  
is still widely applied to viscous (turbulent), cooling
accretion disk models \citep[e.g.][]{gammie01,cossins09, kimura12}, 
despite the mismatch in the included underlying physics. 
We show that including non-ideal physics, such as cooling and
viscosity, in fact modifies the classic dispersion 
relation, and hence the condition for GI. Thus the Toomre condition
alone is insufficient to assess GI in realistic disks.

The goal of this work is to generalize the analytic treatment of disk
GI, by including cooling and viscosity, to allow 
a self-consistent discussion of GI in realistic accretion disk
models.  
%Specifically, we will examine how  cooling and viscosity can
%lead to GI even when $Q>1$, by removing pressure and rotational
%support, respectively.  
We first review in \S\ref{cool_gi} and \S\ref{visc_gi}, respectively,  
how cooling or viscosity can lead to GI even when $Q>1$, by removing
pressure or rotational support. We then 
discuss in \S\ref{frag_intro} how these effects may relate   
to the transition from self-regulated GI to fragmentation seen in
numerical simulations. The rest of this paper is laid out in
\S\ref{plan}.

%{\bf add sentence here more like s 1.4 laying
%out what you are about to talk about. at the very least a transition sentence or two?}

\subsection{Cooling-driven gravitational instability}\label{cool_gi}
Cooling reduces pressure support against self-gravity. The
cooling time  
$\tcool$ is the timescale over which the disk temperature $T$ is relaxed
to some floor value, which may be zero. 
%That is, $\p_t T =
%-(T-T_0)/\tcool$. 
 It is often written as 
\begin{align} \label{beta_def}
  \tcool = \beta \Omega^{-1},
\end{align}
where $\beta$ is the corresponding dimensionless cooling time. This
type of parameterized cooling, first applied by \cite{gammie01},
allows a range of thermodynamic responses to be explored.

In reality, PPD cooling is controlled by radiation   
from dust grains \citep{bell94,dalessio97,chiang97}. Although most
PPDs subject to GI are optically thick, cooling parameterized by the
$\beta$ model formally only captures optically thin cooling when
used in numerical simulations. It is, however, possible to modify the
standard cooling function to mimic optically-thick cooling (see,
e.g. \S\ref{2dppd}).

%%KMK 
Previous work has quantified the role of cooling primarily as a means to reduce the sound speed term
in $Q$ \citep[though see e.g.,][]{clarke07}. Here we will quantify how cooling enables GI
even when $Q>1$. For example, if perturbations can
cool to zero temperature, we find the above dispersion relation is
modified to read 
\begin{align}\label{cool_dr}
s^2 = 2\pi G\Sigma |k| - \kappa^2 - \gamma \left(\frac{\tcool s}{1 +
  \tcool s}\right)c_s^2k^2,
\end{align}
where $\gamma$ is the adiabatic index. Cooling increases the growth
rate by reducing the magnitude of the 
pressure term. In fact, for $t_c\in[0,\infty)$ a formal condition
for instability is  
\begin{align*}
  2\pi G \Sigma |k| > \kappa^2,
\end{align*}
which is what would be obtained from Eq. \ref{classic_dr} with 
pressure neglected ($c_s\to 0$). This condition does not actually
depend on the cooling time, and instability is
possible for any finite $Q$. %at sufficiently small scales 
Thus, the mere presence of cooling changes the qualitative nature of
GI compared to the simple Toomre condition. 
%singular limit

%{\bf numerical implications? if using standard beta cooling, then as
%  long as it's there then sufficiently small scales will be
 % unstbale. can access those scales with high resolution. defer to
 % discussion section}

@@

%turbulence <-> viscosity assumption 
For gaseous accretion disks, a Navier-Stokes viscosity is often
implemented (as above) to mimic hydrodynamic or magneto-hydrodynamic
(MHD) turbulence \citep{shakura73}. How well turbulence can be modeled
as an effective viscosity is a separate issue 
\citep{balbus99}. However, it is reasonable to assume that these
mechanisms, which are observed in simulations  
to transport angular momentum outwards (and mass inwards),
frustate rotational support.

In this work, we use viscosity to model two possible
physical effects of turbulence: heating via dissipation and angular momentum transport. 
%KMK
We emphasize that our model for GI enabled by viscosity (hearafter
viscous GI) does not assume a particular origin for the viscosity. GI
itself or other forms of magnetic or hydrodynamic turbulence are 
allowed in this framework. We will generalize the theory of viscous GI
to include an energy equation  with viscous heating, irradiation,
explicit cooling, as well as three-dimensionality (3D) in order to
consider viscous GI in more realistic PPD models.

@@

\subsection{Relevance to gravito-turbulent disk fragmentation}\label{frag_intro}
We develop a general framework for viscous disks without  
assuming a specific origin of the turbulence that this viscosity 
represents. 
We will, however, apply the theory to `gravito-turbulent'
disks in which the viscosity is associated with some underlying
(classic) GI, as described below.

Consider an initially laminar, $Q\gg1$ disk, without an external
heating mechanism, as it cools. The disk temperature will decline
until $Q=O(1)$, whence non-axisymmetric modes grow and heat the disk
through the dissipation of spiral shocks \citep{cossins09}. 
%new rafikov 16  
This setup permits a quasi-steady, turbulent state with
$Q =O(1)$ in which cooling is balanced by shock heating to maintain
thermal equilibrium \citep{gammie01}. 
If one represents this turbulent heating by a viscous heating, then
the $\alpha$ and $\beta$ parameters defined above will  
be inversely related (see, e.g. Eq. \ref{alpha_beta_relation}).

Numerical experiments, however, show that if the cooling time is too
small (or the viscosity is too large), say,   
\begin{align}\label{frag_cond}
  \beta < \beta_\mathrm{c} \quad
  \left(\alpha>\alpha_\mathrm{c}\right), 
\end{align}
then the disk fragments \citep{gammie01,rice05,rice11}. In fact, in 
the absence of global effects, numerical simulations show that there are
two --- and only two --- possible outcomes for self-gravitating,
cooling disks: gravito-turbulence or fragmentation. 
%more accurately: given a total simulation time, there are two
%outcomes 
%sufficiently long sims -> always frag?

There is considerable debate on the exact value of $\beta_\mathrm{c}$  
and whether or not a critical cooling time can be defined at all
\citep{meru11,lodato11,meru12,paardekooper12,hopkins13}. There are, in
addition, numerical convergence issues when simulating disk  
fragmentation, which we discuss in \S\ref{summary}.

However, it is generally accepted that \emph{steady, gravito-turbulent
  disks do not  
  exist for sufficiently rapid cooling or large viscosity}.  
This is intriguing because, as highlighted in
\S\ref{cool_gi}---\ref{visc_gi}, cooling or viscosity can reduce 
gravitational stability independently of their influence on the exact
value of the classic $Q$. This motivates a physical      
interpretation of disk fragmentation as the inability to maintain a
gravito-turbulent state due to 
secondary instabilities driven by cooling and/or viscosity.

Indeed, we find that the region where a gravito-turbulent PPD is
dynamically unstable agrees with the current consensus on  where, and
the conditions under which, a PPD is expected to fragment based on
Eq. \ref{frag_cond}, with typical critical values cited in the
literature \citep[e.g. $\beta_c\sim
  3,\,\alpha_c\sim0.1$,][]{gammie01,rice05}.

\subsection{Plan}\label{plan}
%plan of this paper 
%aim to analyse GI in realistic disks with minimal input from numerical
%simulation

### 2016-01-28 paper/results2d.tex
revision 3 by MKL

Eq. \ref{inviscid} is a cubic equation in $S$. The 
Routh-Hurwitz criteria imply that stability is ensured if 
\begin{align}\label{stable_condition}
  \gamma > \theta \quad \text{and} \quad 
  Q > \frac{1}{\sqrt{2\theta(2-q)}} 
\end{align}
are both satisfied. 
%gamma/theta condition is related to some kind of thermal instability
%oscilitory convection?? 
%second condition concerning Q is associated with self-g 
Notice these conditions do not actually
depend on the cooling time. 
The disk becomes more unstable with decreasing $\theta$, 
i.e. as the floor temperature, towards which perturbations are cooled, drops. 
{\bf KMK maybe more words here about apparent trend of $Q \rightarrow \inf$ for small $\theta$
even forward ref. lengthscales below?}

If only real growth rates are considered, then violating the second
condition in Eq. \ref{stable_condition} alone is sufficient for instability. 
%from property of cubic crossing the real axis 
In that case the wavenumbers satisfying
\begin{align}
  \theta K^2 - \frac{2}{Q}|K| + 2(2-q) < 0 
\end{align}
are unstable. The range of unstable wavenumbers
increases with decreasing irradiation $\theta$. For $\theta\ll 1$ this range is 
$(2-q)Q\lesssim|K|\lesssim2/\theta Q$. Without irradiation there is no
upper limit to unstable wavenumbers, which could have important
implications for numerical simulations probing large wavenumbers and small scales
at high resolution (see \S\ref{prev_works}).

Consider the most unstable wavenumber $|K_*|$ at which $\p S/\p|K| =
0$ and $S = S_*$. By differentiating Eq. \ref{inviscid}, we obtain 
\begin{align}\label{kstar}
%  |K_*| = \frac{1}{\gamma \qtwo} \left(1 + \frac{1}{\beta S_*}\right). 
  |K_*| = \frac{1+\beta S_*}{Q\left(\theta + \gamma \beta S_*\right)}
\end{align}
%Eq. \ref{kstar} describes how the most unstable  mode is affected by
%cooling ($\beta$) and external heating ($\theta$). 
Inserting this into Eq. \ref{inviscid}, we find the maximum growth
rate satisfies
\begin{align}\label{inviscid_max}
%  S_*^3 + \left[2(2-q)  - \frac{1}{\gamma \qtwo^2}\right]S_* - \frac{1}{\gamma
%    \qtwo^2 \beta} = 0.
  S_*^2 = \frac{1+\beta S_*}{Q^2\left(\theta + \beta\gamma S_*\right)}
  - 2(2-q).
\end{align}
Eq. \ref{kstar}---\ref{inviscid_max} imply $\p_\beta|K_*|,\,\p_\beta
S_*<0 $ for $\gamma>\theta$. That is, as cooling becomes more rapid
the disk becomes more unstable at shorter wavelengths.  {\bf KMK -- I don't follow}
%show dK*/dbeta < 0 
%by diff K* wrt beta and diff S* wrt beta from the equations
%and elim dS*/d beta

>>>>

Eq. \ref{inviscid} is a cubic equation in $S$. The 
Routh-Hurwitz criteria imply that stability is ensured if 
\begin{align}\label{stable_condition}
  \gamma > \theta \quad \text{and} \quad 
  Q > \frac{1}{\sqrt{2\theta(2-q)}} 
\end{align}
are both satisfied. 
%gamma/theta condition is related to some kind of thermal instability
%oscilitory convection?? 
%second condition concerning Q is associated with self-g 
A third criterion, $2(2-q)\gamma Q^2>1$, is formally required, but this
is implied by Eq. \ref{stable_condition}. 
Notice these conditions do not actually depend on the cooling time.

%As perturbations are cooled to lower temperatures
%(decreasing $\theta$), the minimum $Q$ required for stability
%increases.

If we define $Q_\mathrm{eff}=\sqrt{\theta}Q$, the second condition in
Eq. \ref{stable_condition} is $Q_\mathrm{eff}>1$ (for
the Keplerian case). In other words, if one  
insists on applying a Toomre condition to a cooling disk, then the 
Toomre parameter should be calculated with the floor
temperature. %rather than the actual temperature 
At fixed $Q$ this stability condition will fail for sufficiently small 
$\theta$, i.e. as perturbations are allowed to cool to lower
temperatures.

%{\bf KMK maybe more words here about apparent trend of $Q \rightarrow \inf$ for small $\theta$
%even forward ref. lengthscales below?}

If only real growth rates are considered, then violating the second
condition in Eq. \ref{stable_condition} alone is sufficient for instability. 
%from property of cubic crossing the real axis 
In that case the wavenumbers satisfying
\begin{align}
  \theta K^2 - \frac{2}{Q}|K| + 2(2-q) < 0 
\end{align}
are unstable. The range of unstable wavenumbers
increases with decreasing irradiation $\theta$. For $\theta\ll 1$ this range is 
$(2-q)Q\lesssim|K|\lesssim2/\theta Q$. Without irradiation there is no
upper limit to unstable wavenumbers, which could have implications for
numerical simulations probing large wavenumbers and small scales 
at high resolution (see \S\ref{prev_works}).

Consider the most unstable wavenumber $|K_*|$ at which $\p S/\p|K| =
0$ and $S = S_*$. By differentiating Eq. \ref{inviscid}, we obtain 
\begin{align}\label{kstar}
%  |K_*| = \frac{1}{\gamma \qtwo} \left(1 + \frac{1}{\beta S_*}\right). 
  |K_*| = \frac{1+\beta S_*}{Q\left(\theta + \gamma \beta S_*\right)}
\end{align}
%Eq. \ref{kstar} describes how the most unstable  mode is affected by
%cooling ($\beta$) and external heating ($\theta$). 
Inserting this into Eq. \ref{inviscid}, we find the maximum growth
rate satisfies
\begin{align}\label{inviscid_max}
%  S_*^3 + \left[2(2-q)  - \frac{1}{\gamma \qtwo^2}\right]S_* - \frac{1}{\gamma
%    \qtwo^2 \beta} = 0.
  S_*^2 = \frac{1+\beta S_*}{Q^2\left(\theta + \beta\gamma S_*\right)}
  - 2(2-q).
\end{align}
Eq. \ref{kstar}---\ref{inviscid_max} imply $\p_\beta|K_*|,\,\p_\beta
S_*<0 $ for $\gamma>\theta$. As cooling becomes more rapid the maximum 
growth rate increases, and the most unstable wavelength decreases. 
%show dK*/dbeta < 0 
%by diff K* wrt beta and diff S* wrt beta from the equations
%and elim dS*/d beta

If $|S_*|\ll1$, Eq. \ref{inviscid_max} gives the simple solution 
\begin{align}
 \beta S_*\simeq \frac{1-2(2-q)Q^2\theta}{2(2-q)Q^2\gamma-1},
\end{align}
implying that the system can still be formally unstable for arbitarily 
long cooling times. {\bf optional equation?}
